---
title: "Comparing glm on raw data vs glm with splines based on pdp"
author: "Krystian Igras"
date: "29 czerwca 2018"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("breakDown")
library("randomForest")
library("DALEX")
library("xspliner")
library("dplyr")
```

Preparing data and logit
```{r}
data <- HR_data
logit <- function(x) exp(x)/(1+exp(x))
```

Random forest for prediction of `left` variable
```{r}
HR_rf_model <- randomForest(factor(left)~., data = data, ntree = 100)
```

Preparing model with `DALEX::explain`
```{r}
explainer_rf  <- explain(HR_rf_model, data = data, y = data$left,
                         predict_function = function(model, newdata, ...)
                         predict(model, newdata, type = "prob")[,2])

```

Generalized linear model based on raw data
```{r}
HR_glm_model <- glm(left~., data = data, family = "binomial")
```

Generalized linear model with transformed predictors
```{r}
HR_spline_model <- build_spline_model(explainer_rf, data, "left")
```

What exactly `build_spline_model` do?

1. Iteration across all predictors:
  - Calculate pdp curve based on built random forest (`explainer_rf`) 
  - Approximate pdp with spline (here using `mgcv` package)
  - When variable is factor, ordered or has not enough unique values there is used identity (`I`) instead of spline
2. Building glm model when each predictor is transformed with above function (identity or spline)  
  
Let's compare two models:

# AIC:
```{r}
HR_glm_model$aic
HR_spline_model$aic
```

Based on AIC "spline" model is much better.

# Accuracy:
```{r}
data$pred_glm <- round(logit(predict(HR_glm_model, data)))
data$pred_spline <- round(logit(predict(HR_spline_model, data)))
sum(data$pred_glm == data$left) / nrow(data)
sum(data$pred_spline == data$left) / nrow(data)
```

We can also see that spline model has higher accuracy.
