---
title: "Methods and xspliner environment"
author: "Krystian Igras"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methods and xspliner environment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Predict

As xspliner final model is GLM, predict method is just wrapper of `stats::predict.glm` function. 
Let's see it on the below example:
```{r message = FALSE}
library(xspliner)
library(randomForest)
library(magrittr)

rf_iris <- randomForest(Petal.Width ~  Sepal.Length + Petal.Length + Species, data = iris)
model_xs <- xspline(Petal.Width ~ 
  Sepal.Length + 
  xs(Petal.Length, effect = list(grid.resolution = 100), transition = list(bs = "cr")) + 
  xf(Species, transition = list(stat = "loglikelihood", value = -300)),
  model = rf_iris)
newdata <- data.frame(
  Sepal.Length = 10, 
  Petal.Length = 2, 
  Species = factor("virginica", levels = levels(iris$Species)))
predict(model_xs, newdata = newdata)
```

# Summary

Summary method allows you to check the basic model details. See below what possibilities the method to xspliner model offers.

## GLM summary
Standard summary method is just wrapper for `summary::glm`. In order to use this just type:
```{r}
summary(model_xs)
```

## Predictor based summary
Summary method allows you to check details about transformation of specific variable.

Standard usage `summary(xspliner_object, variable_name)`

*Quantitative variable*
When predictor is quantitative variable its transition is based on GAM model. For this case summary displays summary of that model.

```{r}
summary(model_xs, "Petal.Length")
```

*Qualitative variable*

In case of qualitative predictor, the method displays data.frame storing information how factors were merged during the transition.

```{r}
summary(model_xs, "Species")
```

## Surrogate vs Black Box comparison
Providing model parameter instead of predictor, the summary displays a few statistics that compares original model with surrogate one.
All statistics definitions are included in summary.xspline documentation.

Here we show one example for classification model.

For this example we will use graduation data (source: ) and build svm model as black box.
```{r}
library(xspliner)
library(e1071)
set.seed(1)
data <- read.csv("binary.csv")
data$admit <- as.factor(data$admit)
# Classification model
grad.svm <- svm(admit ~ ., data = data, probability = TRUE)
grad.xs <- xspline(admit ~ xs(gre) + xs(gpa) + rank, model = grad.svm)

# Comparing summaries requires providing prediction function
# Prediction as probability for success
prob_svm <- function(object, newdata) attr(predict(object, newdata = newdata, probability = TRUE), "probabilities")[, 2]
prob_xs <- function(object, newdata) predict(object, newdata = newdata, type = "response")
summary(grad.xs, model = grad.svm, newdata = data, prediction_funs = list(prob_xs, prob_svm))
# Prediction as final category
response_svm <- function(object, newdata) predict(object, newdata = newdata)
response_xs <- function(object, newdata) {
  y_levels <- levels(newdata[[environment(object)$response]])
  factor(y_levels[(predict.glm(object, newdata = newdata, type = "link") > 0) + 1], levels = y_levels)
}
summary(grad.xs, model = grad.svm, newdata = data, prediction_funs = list(response_xs, response_svm))
```

# Print

Print method works similarly to the summary. In case of passing just the model, standard `print.glm` is used.

```{r}
print(model_xs)
```

## Predictor based print
Summary method allows you to check details about transformation of specific variable.

Standard usage `print(xspliner_object, variable_name)`

*Quantitative variable*
When predictor is the quantitative variable its transition is based on GAM model. For this case print uses standard `print.gam` method.

```{r}
print(model_xs, "Petal.Length")
```

*Qualitative variable*

In case of qualitative predictor, standard `print.factorMerger` method is used.

```{r}
print(model_xs, "Species")
```

# Plot

You can see all details in [graphics](./graphics.html)

# Transition

Transition method allows you to extract objects used during building transition of variables.
There are three possible object types that can be extracted.

## Extracting effect

Each transition is built on top of the black box response data. For example the default response for quantitative variables is PDP - for qualitative ones ICE.

In order to extract the effect use transition method with `type` parameter equals to data

```{r}
transition(model_xs, predictor = "Petal.Length", type = "data") %>% 
  head

transition(model_xs, predictor = "Species", type = "data") %>% 
  head
```

## Extracting transition model

After we built transition basing on continuity of variable specific model is created.
In case of quantitative predictor we build GAM model in order to get spline approximation of effect.
In case of qualitative predictor we build factorMerger object and get optimal factor division on that.

To extract the model, use transition method with `type = "base"`:

```{r}
transition(model_xs, predictor = "Petal.Length", type = "base")

transition(model_xs, predictor = "Species", type = "base")
```

## Extracting transition function

The final result of building transition is transformation function, that is used in the final GLM model estimation.

To extract the function just use transition method with `type = "function"`.

```{r}
petal_length_xs <- transition(model_xs, predictor = "Petal.Length", type = "function")
x <- seq(1, 7, length.out = 50)
plot(x, petal_length_xs(x))
```

```{r}
species_xf <- transition(model_xs, predictor = "Species", type = "function")
species_xf(c("setosa", "versicolor", "virginica"))
```
